Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-07-27 TEX/draft_biology.tex
morning session

\newcommand{\LL}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\ep}{\varepsilon}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\GCD}{\mathbf{GCD}}
\newcommand{\XX}{\mathcal{X}}
\newcommand{\SUM}{\text{sum}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\kk}{\textbf{k}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\mb}{\mathbf}
\newcommand{\mk}{\mathfrak}

@@

\newcommand{\reals}{\mathbb{R}}
\newcommand{\RL}{\mathbb{R}^L}
\newcommand{\CL}{\mathbb{C}^L}
\newcommand{\RN}{\mathbb{R}^N}
\newcommand{\RNN}{\mathbb{R}^{N\times N}}
\newcommand{\CNN}{\mathbb{C}^{N\times N}}
\newcommand{\inner}[1]{\left\langle {#1} \right\rangle}
\newcommand{\hx}{\hat{x}} 
\newcommand{\one}{\mathbf{1}} 
\newcommand{\SNR}{\ensuremath{\textsf{SNR}}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}

@@

The techniques we advocate allow to recover a signal hidden in noisy micrographs without (even implicitly) detecting the location of signal repetitions in these micrographs. To illustrate the underlying principles of the method, we start with a toy model where the target signal is an unknown 2-D image. The latter appears several times in noisy micrographs, at unknown locations. Toward recovery, we first compute the second-order autocorrelation of the micrographs. We then establish that the autocorrelation of the target image can be recovered from the autocorrelation of the micrographs in a straightforward manner. The more micrographs we collect, the more accurate the estimate of the image's autocorrelation. 
The target image is then estimated from its autocorrelation using a standard inversion algorithm (up to elementary symmetries).
The same principles carry over to the more challenging task of 3-D reconstruction from micrographs.
The Methods Section provides additional details.

@@

We compute the average second-order autocorrelation of the micrographs. This is a particularly simple computation which can be efficiently executed with a fast Fourier transform on GPUs. In the Methods Section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the autocorrelation of the unknown image itself. Mathematically, it can be shown that for any noise level, the quality of this estimate improves steadily as the amount of data grows~\cite{bendory2018estimation}. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}, initialized randomly.

@@

\subsection{2-D experiment (we need a name)}
To explain the method, we begin by deriving the algebraic relation between the autocorrelation functions of the micrographs and the autocorrelation functions of the planted image.
%We derive algebraic relations between the autocorrelation functions of the micrographs and the autocorrelation functions of the target signals. For ease of exposition, we do so in the 1-D case. Extension to the 2-D case is straightforward. We also give additional technical details regarding the two experiments presented in Section~\ref{sec:results}.

Let $x$ be the sought image and let $y\in\RNN$ be the observed micrograph (notice that it is equivalent to think of the data as being one long micrograph or multiple smaller micrographs concatenated into one.)
The forward model (or ``image'' formation model) is as follows. An unknown binary signal  indicates (with 1's) the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise:
\begin{align}
	y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2 I_N),
	\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution. Let $\textbf{i}:=(i_1,i_2)$ and similarly for \textbf{j}.
The binary signals obey the following property: \TODO{Rewrite in terms of $\delta$ functions for consistency with 3D}
\begin{align}
	\textrm{If } s[\textbf{i}] = 1 \textrm{ then } s[\textbf{j}] = 0 \textrm{ for all}  \|\textbf{i}-\textbf{j}\|_\infty\leq 2L-1.
	\label{eq:spacing}
\end{align}
In words: the starting positions of any two occurrences must be separated by at least $2L-1$ positions in each direction, so that their end points are necessarily separated by at least $L-1$ signal-free entries in the micrograph.

From $y$, we aim to recover $x$. In this simplified model, we assume to know the number of  occurrences of the signal. In contrast, particle picking is the task of estimating the binary signal $s $, which cannot  be performed reliably if $\sigma$ is large (that is, at low $\SNR$.)

%We denote the set of these nonzero values by 
% $\mathcal{S}_i$ and its cardinality by $\vert \mathcal{S}_i\vert = M_i$. 
%By assuming that all $\mathcal{S}_i$'s are disjoint,  we let $s = \sum_{i=1}^Ks_i$, $\mathcal{S} = \bigcup_{i=1}^{K} \mathcal{S}_i$ and  $\vert \mathcal{S}\vert :=M =  \sum_{i=1}^{K}M_i$.  %Neither the $M_i$'s nor $M$ are assumed to be known.

%
%Considering~\eqref{eq:model}, one can think of this inverse problem as a mixture of blind deconvolution problems between binary signals and the target signals. Related literature is briefly surveyed in Appendix~\ref{sec:related_literature}.

%In order to estimate the mixture of autocorrelations, we assume that the support of $s$ is not clustered. In particular, we assume that
%\begin{align}
%  \vert i-j \vert\geq 2L-1,   \quad \text{for all } i,j\in\mathcal{S} \text{ such that } i\neq j.
%  \label{eq:spacing}
%\end{align}
%The goal is to estimate $x_1,\ldots,x_K$ from $y$.

@@

\paragraph{Aperiodic autocorrelation functions}
%\subsection{Aperiodic autocorrelation functions} \label{sec:aperiodic_ac}

%\TODO{$L$ is a bad notation here since it could be L or n; and k is poor notation too because we use it to sum up to K.}
%
%In general, for a signal $z$ of length $m$, the autocorrelation of order $q \geq 1$ is given for any integer shifts $\ell_1, \ldots, \ell_{q-1}$ by
%\begin{align}
%	a_z^q[\ell_1,\ldots,\ell_{k-1}]  & = \frac{1}{m} \sum_{i=-\infty}^{+\infty} z[i]z[i+\ell_1]\cdots z[i+\ell_{q-1}],
%	\label{eq:ac_general}
%\end{align}
%where indexing of $z$ out of the range $0, \ldots, m-1$ is zero-padded.
%For our purposes, this will be applied both to $x_k$'s (each of length $L$) and to $y$ (of length $N$).
For the purpose of this experiment, we need only the mean of the signal and its second-order autocorrelation defined for an arbitrary signal $z\in\mathbb{R}^{m\times m}$ by \TODO{to verify}
 \begin{align} 
 a_z^2[\ell_1,\ell_2] & = \frac{1}{m^2} \sum_{i_1 = \max\{0, -\ell_1\}}^{m-1 + \min\{0, -\ell_1\}} \sum_{i_2 = \max\{0, -\ell_2\}}^{m-1 + \min\{0, -\ell_2\}}z[i_1,i_2]z[i_1+\ell_1,i_2+\ell_2]. 
 \end{align}
 It can be shown that in the limit of $N\to\infty$,  the mean of the microgaph is equal to the mean of the signal $x$, scaled by the scalar 
 \begin{align}
 \gamma & = \frac{M L^2}{N^2},
 \end{align}
 which denotes the density of $x$ in $y$ (that is, the fraction of entries of $y$ occupied by occurrences of $x$.) The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$.
 Similarly, the second-order autocorrelations are related through 
 \begin{align*}
	\lim_{N\to\infty} a_y^2[\ell_1,\ell_2] & = \gamma a_{x}^2[\ell_1,\ell_2] + \sigma^2\delta[\ell_1,\ell_2].
 \end{align*}
 These relations, and their extension to multiple planted images, are proven and discussed in~\cite{bendory2018estimation}.
%Explicitly, the first-, second- and third-order autocorrelations are given by
%\begin{align} 
%	a_z^1 & = \frac{1}{m} \sum_{i=0}^{m-1} z[i], \nonumber\\
%	a_z^2[\ell] & = \frac{1}{m} \sum_{i = \max\{0, -\ell\}}^{m-1 + \min\{0, -\ell\}} z[i]z[i+\ell], \nonumber\\
%	a_z^3[\ell_1,\ell_2] & = \frac{1}{m} \sum_{i = \max\{0, -\ell_1, -\ell_2\}}^{m-1 + \min\{0, -\ell_1, -\ell_2\}} z[i]z[i+\ell_1]z[i+\ell_2]. \label{eq:ac_special}
%\end{align}
%The autocorrelation functions have symmetries. Specifically, $a_z^2[\ell] = a_z^2[-\ell]$, and
%\begin{align*}
%	a_z^3[\ell_1,\ell_2] = a_z^3[\ell_2,\ell_1]=a_z^3[-\ell_1,\ell_2-\ell_1].
%\end{align*}
%%Taking these symmetries into consideration, one can show that $a_z^1, a_z^2$ and $a_z^3$ contain \TODO{TKTK--I think it was commented in the latex} (respectively) non-trivial distinct real numbers, generically.
%\paragraph{Estimating the autocorrelation function of a single signal.}
%For the special case $K = 1$ where a single signal $x = x_1$ must be recovered, the relation between autocorrelations of the micrograph and those of $x$ is particularly simple, so that we treat it first. It is useful to introduce some notation: let $M$ denote the number of occurrences of $x$ in $y$, and let
%\begin{align}
%	\gamma & = \frac{M L}{N}
%%\end{align}
%denote the density of $x$ in $y$ (that is, the fraction of entries of $y$ occupied by occurrences of $x$.) The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$.
%One simple observation is that the first-order autocorrelation of $y$ (its mean) is independent of the locations of $x$. Since the noise is independent of the signal, the mathematical expectation of $a_y^1$ is easily seen to be:\footnote{We did not fully specify a random generating model for the location vector $s$. The expectation is still well defined specifically because the quantity under consideration is independent of $s$ under the assumptions.}
%\begin{align*}
%	\E\{ a_y^1 \} & = \gamma a_x^1.
%\end{align*}
%We consider the asymptotic regime where $M, N\to\infty$, while $\gamma$ remains constant (we see an increasingly large micrograph, containing increasingly many signal occurrences, with constant signal density.) In that regime, the law of large numbers gives meaning to the following statement:
%\begin{align*}
%	\lim_{N\to\infty} a_y^1 & = \gamma a_{x}^1.
%\end{align*}
%Thus, given enough data, if $\gamma$ is known, we can estimate $a_x^1$ from $y$. (We show later how to estimate $\gamma$ as well.)
%The spacing constraint~\eqref{eq:spacing} gives rise to more powerful observations. Consider the second-order autocorrelation in particular: $a_y^2[\ell]$ computes the correlation between $y$ and a copy of $y$ shifted by $\ell$ entries. Considering $\ell$ only in the range $0, \ldots, L-1$, one can see that any given occurrence of $x$ in $y$ is only ever correlated with itself (with the same shift $\ell$), and never with another occurrence. As a result,
%\begin{align*}
%	\lim_{N\to\infty} a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell]
%\end{align*}
%for $\ell = 0, \ldots, L-1$, where $\delta$ denotes the Kronecker delta function. The last part captures the autocorrelation of the noise. Notice that, even if $\sigma$ is unknown, entries $\ell = 1, \ldots, L-1$ still provide useful information about $a_x^2$.
%Along the same lines, one can establish a relation for third-order autocorrelations:
%\begin{align}
%	\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2] + \sigma^2\gamma a_{x}^1 \cdot \big(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]\big),
%	\label{eq:data_ac_k1}
%\end{align}
%for $\ell_1,\ell_2 = 0, \ldots, L-1$. Here too, few entries are affected by $\sigma$ in the limit.
%Detailed derivations for identities in this and the next part are given in Appendix~\ref{sec:autocorrelation_computation}.
%%
%\paragraph{Estimating the autocorrelation function of multiple signals.}
%Returning to the general case $K \geq 1$, let $M_1, \ldots, M_K$ denote the number of occurrences of signals $x_1, \ldots, x_K$ respectively, and define
%\begin{align}
%	\gamma_k & = \frac{M_k L}{N}, & \gamma & = \sum_{k=1}^K\gamma_k.
%\end{align}
%As above, we consider the asymptotic regime where $M_1,\ldots,M_K,N\to\infty$ while preserving the ratios $\gamma_k$ constant.
%% 
%Still under the spacing constraint~\eqref{eq:spacing}, similarly to the developments above, one can estimate a mixture of the autocorrelations of the $K$ target signals from the autocorrelations of the micrograph:
%\begin{align}
%\lim_{N\to\infty} a_y^1 & = \sum_{k=1}^K\gamma_k a_{x_k}^1, \nonumber\\
%\lim_{N\to\infty} a_y^2[\ell] & = \sum_{k=1}^K\gamma_k a_{x_k}^2[\ell] +\sigma^2\delta[\ell],  \label{eq:data_ac}\\
%\lim_{N\to\infty} a_y^3[\ell_1,\ell_2] & = \sum_{k=1}^K\gamma_k a_{x_k}^3[\ell_1,\ell_2] + \sigma^2\left(\sum_{k=1}^K\gamma_k a_{x_k}^1\right)(\delta[\ell_1,0]+\delta[0,\ell_2]+\delta[\ell_1,\ell_2]), \nonumber
%\end{align}
%where $\ell, \ell_1, \ell_2 = 0, \ldots, L-1$. The left hand side is straightforward to estimate from data: it provides a succinct summary of it. The right hand side involves polynomial functions of unknowns $\gamma_1, \ldots, \gamma_K, x_1, \ldots, x_K$, and possibly $\sigma^2$. The task is to solve these polynomial equations in a robust way.

%\TODO{It would be more natural to present the 1D XP first given everything that preceeds, then to present the 2D XP, where we can explain it's a pretty special case.}
%\paragraph{Numerical experiment with three 1-D signals.}
%For the 1-D experiment depicted in Figure~\ref{fig:1Dheterosignals}, we fix $K = 3$ signals of length $L = 21$. Following the forward model described at the beginning of this section, we generate an observation $y$ of length $12.3 \cdot 10^9$. Each of the three signals appears, respectively (and approximately), $30.0 \cdot 10^6$, $20.0 \cdot 10^6$ and $10.0 \cdot 10^6$ times in $y$ for a total of exactly $60 \cdot 10^6$ occurrences, such that at least $L-1$ zeros separate any two occurrences of any signals. 
%This is done by randomly selecting $60 \cdot 10^6$ placements in $y$, one at a time with an accept/reject rule based on the separation constraint and locations picked so far. For each placement, one of the three signals is picked at random according to the proportions $1/2, 1/3, 1/6$. Then, i.i.d.\ Gaussian noise with mean zero and standard deviation $\sigma = 3$ is added, to form the observed $y$. The resulting $\SNR$ of $y$
%% sqrt((m_want*sum(X.^2)')/(sigma^2*n))
%is about 1/9.
%This is enough noise to make cross-correlations of $y$ even with the true signals display peaks at essentially random locations, uninformative of the actual locations of the signal occurrences. Thus, we contend that it would be difficult for any algorithm to locate the signal occurrences, let alone to classify them according to which signal appears where.
%%\TODO{TB: I would place the equation counting argument in the theory section in the paragraph of open questions (I marked th place)}
%%
%Given the observation $y$, we proceed to compute the autocorrelations. The first-order autocorrelation is straightforward. For second-order autocorrelations, notice from equation~\eqref{eq:data_ac} that $a_y^2[\ell]$ suffers no noise-induced bias for $\ell$ in $1$ to $L-1$. Thus, we omit $\ell = 0$, which has the practical effect that we need not know $\sigma$ to make sense of the computed quantities. Likewise, for third-order autocorrelations, $a_y^3[\ell_1, \ell_2]$ for $0 \leq \ell_1, \ell_2 \leq L-1$ such that $\ell_2 \leq \ell_1$ includes all relevant entries for our purpose (this accounts for symmetries), and we further exclude any such that $\ell_1, \ell_2$ or $\ell_1 - \ell_2$ are zero to avoid the need to estimate $\sigma$---there are $\frac{(L-1)(L-2)}{2}$ remaining entries. We have
%\begin{align*}
%1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
%\end{align*}
%coefficients in total. Since we aim to estimate $KL$ parameters (for the $K$ signals of length $L$) plus $K$ parameters (for the densities $\gamma_k$), an absolute upper bound on $K$ (simply to ensure we have at least as many equations as we have unknowns) is
%\begin{align*}
%K(L+1) \leq \frac{1}{2} L (L-1) + 1.
%\end{align*}
%Thus, $(L-1)/2$ (up to a small approximation) is an absolute upper limit on $K$. \TODO{We may want to cite MRA literature here.}
%In practice, the autocorrelations are computed on disjoint segments of $y$ of length $100\cdot10^6$ and added up, without correction for the junction points. Segments are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.
%Having computed the moments of interest, we now estimate signals $x_1, \ldots, x_K$ and coefficients $\gamma_1, \ldots, \gamma_K$ which agree with the data. We choose to do so by running an optimization algorithm on the following nonlinear least-squares problem:
%\begin{multline}
%\min_{\substack{\hat x_1, \ldots, \hat x_K \in \reals^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2,
%\label{eq:optim1D}
%\end{multline}
%where $W \geq L$ is the length of the sought signals and the weights are set to $w_1 = 1/2, w_2 = 1/2n_2, w_3 = 1/2n_3$, where $n_2, n_3$ are the number of moments used: $n_2 = L-1$, $n_3 = \frac{(L-1)(L-2)}{2}$ (weights could also be set in accordance with variance estimates as in~\cite{boumal2017heterogeneous}).
%Setting $W = L$ (as is a priori desired) is problematic because the above optimization problems appears to have numerous poor local optimizers.
%%Since we can only initialize randomly at first, this approach would often fail in practice. Alternatively,
%Thus, we first run the optimization with $W = 2L-1$. This problem appears to have few poor local optima, perhaps because the additional degrees of freedom allow for more escape directions. Since we hope the signals estimated this way correspond to the true signals zero-padded to length $W$, we extract from each one a subsignal of length $L$ that has largest $\ell_2$-norm. This estimator is then used as initial iterate for~\eqref{eq:optim1D}, this time with $W = L$. We find that this procedure is reliable for a wide range of experimental parameters. To solve~\eqref{eq:optim1D}, we run the trust-region method implemented in Manopt~\cite{manopt}, which allows to treat the positivity constraints on coefficients $\hat \gamma_k$. Notice that the cost function is a polynomial in the variables, so that it is straightforward to compute it and its derivatives.
%%\TODO{Should we do variable projection for the gammas, that is, exploit the fact the problem is a regular least squares in the gammas (up to the positivity constraints) to substitute the explicit optimum for them? Not sure it's worth the effort. -- Ok, it's probably not a good idea, because even with fixed gammas to the correct value, optimization takes a while.}
%%\TODO{Do we still need to stress at this point that the optimization part has complexity independent of length of observation? Should be pretty clear at this point already.}
%%
%%
%%\TODO{separation}
%%
%%For the 1-D experiment, we worked with three signals of length $L = 21$ and generated the data in the same way as in the 2-D example.
%%The only difference is that here, for each
%%placement, one of the three signals is picked at random proportionally to the desired number
%%of occurrences of each.
%%In this experiment, we computed the first three autocorrelation functions. 
%%We do not assume to know the number of occurrences of each signal $\gamma_i$ ahead and we removed the biased terms (see~\eqref{eq:data_ac_k1}) so we do not need to know $\sigma$ either. 
%%\TODO{In Appendix~\ref{sec:theory} we provide an argument on the number of equations we get from the first three autocorrelation functions. -- Was this removed? I can't find it.}
%%To estimate the signal, we use an optimization algorithm on the following nonlinear least-squares problem that estimates the signals and their density of occurrences simultaneously:
%%\begin{multline}
%%\min_{\substack{\hat x_1, \ldots, \hat x_K \in \reals^{W} \\ \hat \gamma_1, \ldots, \hat \gamma_K > 0}} w_1 \left( a_y^1 - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^1 \right)^2 + w_2 \sum_{\ell = 1}^{L-1} \left( a_y^2[\ell] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^2[\ell] \right)^2 + \\ w_3 \sum_{\substack{2\leq\ell_1\leq L-1 \\ 1 \leq \ell_2 \leq \ell_1-1}} \left( a_y^3[\ell_1, \ell_2] - \sum_{k=1}^K \hat \gamma_k a_{\hat x_k}^3[\ell_1,\ell_2] \right)^2.
%%\label{eq:optim1D}
%%\end{multline}
%%where $W \geq L$ is the length of the sought signals and $w_1 = \frac{1}{2}, w_2 =\frac{1}{2}(L-1), w_3 = \frac{1}{(L-1)(L-2)}$; see~\cite{boumal2017heterogeneous} see for discussion on how to choose the weights properly.
%%Setting $W = L$ (as is a priori desired) is problematic because the above optimization problems appears to have numerous poor local optimizers.
%%Since we can only initialize randomly at first, this approach would often fail in practice. Alternatively,
%%Thus, we first run the optimization with $W = 2L-1$. This problem appears to have fewer poor local optima, perhaps because the additional degrees of freedom allow for more escape directions. Since we hope the signals estimated this way correspond to the true signals zero-padded to length $W$, we extract from each one a subsignal of length $L$ whose autocorrelation functions are closest to the measured ones in the sense of~\eqref{eq:optim1D}. 
%%This estimator is then used as initial iterate for~\eqref{eq:optim1D}, this time with $W = L$. We find that this procedure is reliable for a wide range of experimental parameters. To solve~\eqref{eq:optim1D}, we run the trust-region method implemented in Manopt~\cite{manopt}, which allows to treat the positivity constraints \TODO{reference} on coefficients $\hat \gamma_k$. Notice that the cost function is a polynomial in the variables, so that it is straightforward to compute it and its derivatives.

\paragraph{Numerical experiment with 2-D image.}

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $P$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an $\SNR$ of approximately $1/20$.
An example of a micrograph's excerpt is presented in the right panel of Figure~\ref{fig:micro_example}.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

In this experiment, we assume we know the noise level $\sigma$ and the total number of occurrences of the target image across all micrographs.
In stark contrast with the 1-D setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin~\cite{hayes1982reconstruction} (see also~\cite{bendory2017fourier} for a review). This is because the second-order autocorrelations correspond to the Fourier magnitudes of the signal through the 2-D Fourier transform. 
Therefore, we estimate the signal's Fourier magnitudes (or power spectrum) from the Fourier magnitudes of the micrographs, at the cost of one 2-D fast Fourier transform (FFT) per micrograph. These can be computed highly efficiently and in parallel.

To recover the target image from the estimated power spectrum, we use a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
\begin{align*}
	z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z))
\end{align*}
on an image $z$ of size $2L\times 2L$.
We set the parameter $\beta$ to 1.
The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, possibly reflected through its origin, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner.

In order to compare the performance in multiple cases and at different noise levels, the algorithm is stopped after a fixed number of iterations (1000) and the iterate with the smallest error compared to the ground truth (up to the reflection ambiguity) is chosen as the solution. While this cannot be done in practice (since we do not have access to the ground truth to determine which iterate is best), this procedure enables us to compare a large number of instances in different noise environments. \TODO{Note the last two sentences!}

>>>>

\newcommand{\LL}{\mathcal{L}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\ep}{\varepsilon}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\GCD}{\mathbf{GCD}}
\newcommand{\XX}{\mathcal{X}}
\newcommand{\SUM}{\text{sum}}
\newcommand{\1}{\mathbf{1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\ii}{\textbf{i}}
\newcommand{\jj}{\textbf{j}}

\newcommand{\II}{\mathcal{I}}
\newcommand{\kk}{\textbf{k}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\mb}{\mathbf}
\newcommand{\mk}{\mathfrak}
\newcommand*\Bell{\ensuremath{\boldsymbol\ell}}

@@

\newcommand{\reals}{\mathbb{R}}
\newcommand{\RL}{\mathbb{R}^L}
\newcommand{\tamir}{x}
\newcommand{\CL}{\mathbb{C}^L}
\newcommand{\RN}{\mathbb{R}^N}
\newcommand{\RNN}{\mathbb{R}^{N\times N}}
\newcommand{\RPP}{\mathbb{R}^{P\times P}}
\newcommand{\CNN}{\mathbb{C}^{N\times N}}
\newcommand{\inner}[1]{\left\langle {#1} \right\rangle}
\newcommand{\hx}{\hat{x}} 
\newcommand{\one}{\mathbf{1}} 
\newcommand{\SNR}{\ensuremath{\textsf{SNR}}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}

@@

The techniques we advocate allow to recover a signal hidden in noisy micrographs without (even implicitly) detecting the location of signal repetitions in these micrographs. To illustrate the underlying principles of the method, it is instructive to start with a toy model where the target signal is an unknown 2-D image. The latter appears several times in noisy micrographs, at unknown locations. Toward recovery, we first compute the second-order autocorrelation of the micrographs. We then establish that the autocorrelation of the target image can be recovered from the autocorrelation of the micrographs in a straightforward manner. The more micrographs we collect, the more accurate the estimate of the image's autocorrelation. 
The target image is then estimated from its autocorrelation using a standard inversion algorithm (up to elementary symmetries).
The same principles carry over to the more challenging task of 3-D reconstruction from micrographs.
The Methods Section provides additional details.

@@

We compute the average second-order autocorrelation of the micrographs. This is a particularly simple computation which can be efficiently executed with a fast Fourier transform (FFT) on GPUs. In the Methods Section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the autocorrelation of the unknown image itself. Mathematically, it can be shown that for any noise level, the quality of this estimate improves steadily as the amount of data grows~\cite{bendory2018estimation}. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}, initialized randomly.

@@

\subsection{Toy 2-D experiment}

Let $x\in\RPP$ be the target image and let $y\in\RNN$ be  one of  $K$ observed micrographs. 
%If more than one micrograph is available, then it is sufficient to average the autocorrelations we compute below over the micrographs.
Independently for each micrograph, the image formation model is as follows. An unknown binary signal $s\in\{0,1\}^{(N-P)\times(N-P)}$ indicates the starting positions of all occurrences of $x$ in $y$, so that, with additive white Gaussian noise, we observe:
\begin{align}
	y & =  x \ast s + \varepsilon, & \varepsilon & \sim \mathcal{N}(0,\sigma^2I),
	\label{eq:model}
\end{align}
where $\ast$ denotes linear convolution, $\ii:=(i_1,i_2)$ identifies one location in the image and $\ii=(0,0)$ indicates the  upper left corner.
We define the images to be zero outside their support.
The binary signals $s$ obey the following separation property: 
\begin{align}
	\textrm{If } s[\ii] = 1 \textrm{, then } s[\jj] = 0  \textrm{   for all $\jj=(j_1,j_2)$ such that   }  \max\{\vert i_1-j_1\vert, \vert i_2-j_2\vert\}\leq 2P-1.
	\label{eq:spacing}
\end{align}
In words: the upper left corners of any two occurrences of the target image $x$ in a micrograph must be separated by at least $2P-1$ pixels in each direction.
Hence, each image is separated by a signal-free area that will ease the interpretation of the autocorrelation analysis. 
From the micrographs, we aim to  recover $x$ only. In this simplified model, we assume  the total number of  occurrences of the signal across micrographs is known. 
In the 3-D experiment we do not assume knowledge of the number of projections in the micrographs.
%In contrast, particle picking is the task of estimating the binary signal $s $, which cannot  be performed reliably if $\sigma$ is large (that is, at low $\SNR$.)

@@

\paragraph{Aperiodic autocorrelation functions.}
The second-order autocorrelation of  an arbitrary image $z\in\mathbb{R}^{m\times m}$ is defined by
  \begin{align} 
 a_z^2[\Bell] & = \frac{1}{m^2} \sum_{i_1 = \max\{0, -\ell_1\}}^{m-1 + \min\{0, -\ell_1\}} \sum_{i_2 = \max\{0, -\ell_2\}}^{m-1 + \min\{0, -\ell_2\}}z[\ii]z[\ii+\Bell]. 
 \end{align}
 Below, the image $z$ might be either a target image $x$ or the micrograph $y$. 
 It can be shown that in the limit of infinitely many micrographs $K\to\infty$,  the mean pixel value of the micrographs is equal to the mean pixel value of the signal $x$, scaled by  
 \begin{align}
 \gamma & = \frac{M P^2}{N^2},
 \end{align}
 where $M$ is the number of occurrences of $x$ across micrographs. 
 This  is the fraction of entries of all micrographs occupied by occurrences of $x$, that is, the  density of the signal. 
  The spacing constraint~\eqref{eq:spacing} imposes $\gamma\leq\frac{P}{2P-1}\approx 1/2$.

 % Similarly, as $K\to\infty$ the second-order autocorrelations are related through 
 Similarly, numbering micrographs as $y_1, \ldots, y_K$, the empirical average of the second-order autocorrelations of the micrographs converges as follows:
 \begin{align} \label{eq:a2}
	\lim_{K \to \infty} \frac{1}{K} \sum_{k=1}^{K} a_{y_k}^2 & = \gamma a_{x}^2 + \sigma^2\delta,
 \end{align}
 where $\delta[\Bell]=1$ for $\Bell=(0,0)$ and zero otherwise.
 Thus, for any noise level and without the need to even implicitly locate the images $x$ in the micrographs, given sufficiently many micrographs we are able to compute the second-order autocorrelation of $x$ itself. Recovering $x$ from $a_x^2$ can be done up to symmetries, as detailed below.
 Equation~\eqref{eq:a2} is proven and its extension to higher-order autocorrelations and multiple distinct planted images is discussed in~\cite{bendory2018estimation}.

\paragraph{Numerical experiment with a 2-D image.}

For the 2-D experiment shown in Figures~\ref{fig:Einst_example} and~\ref{fig:error_per_micro}, we generate $K$ micrographs of size $4096\times 4096$ pixels. 
In each micrograph, we place Einstein's image (of zero mean) of size $50\times 50$  in random locations, while preserving the separation condition~\eqref{eq:spacing}.  
This is done by randomly selecting $4000$ placements in the micrograph, one at a time with
an accept/reject rule based on the separation constraint and locations picked so far.
On average, $700$ images are placed in each micrograph.   
Then, i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ is added, inducing an $\SNR$ of approximately $1/20$.
An example of a micrograph's excerpt is presented in the right panel of Figure~TKTK.
%Different micrographs are handled sequentially on a GPU, as GPUs are particularly well suited to execute simple instructions across large vectors of data. If multiple GPUs are available, segments can of course be handled in parallel.

In this experiment, we assume $\gamma$ and the noise level $\sigma$ are known. In this setup, the second-order autocorrelation determines almost any target image uniquely, up to reflection through the origin~\cite{hayes1982reconstruction}. The second-order autocorrelation can be computed at the cost of one 2-D FFT per micrograph. These can be computed highly efficiently and in parallel. To recover the image $x$ itself, we use a standard algorithm used for phase retrieval called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}.

%
%To recover the target image from the estimated power spectrum, we use a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR). This algorithm iterates the map
%\begin{align*}
%	z & \leftarrow z + \beta (P_2(2P_1(z) - z) - P_1(z))
%\end{align*}
%on an image $z$ of size $2L\times 2L$.
%We set the parameter $\beta$ to 1.
%The map is designed so that, if the estimated power spectrum is exact, then fixed points contain Einstein's image in the upper-left corner of size $L \times L$, possibly reflected through its origin, and zeros elsewhere. The operator $P_2(z)$ combines the Fourier phases of  the current estimation $z$ with the estimated Fourier magnitudes. The operator $P_1(z)$ zeros out all entries of $z$ outside the $L\times L$ upper-left corner.

In order to compare the performance in multiple cases and at different noise levels, the RRR is stopped after a fixed number of iterations (1000)\TODO{check} and the iterate with the smallest error compared to the ground truth (up to the reflection ambiguity) is chosen as the solution. 
While this cannot be done in practice (since we do not have access to the ground truth to determine which iterate is best), this procedure enables us to compare a large number of instances in different noise environments. \TODO{Note the last two sentences!}
\TODO{Let's try with the last iterate and see}
